Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00

### 2019-03-02 manuscript.Rmd
text updates

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across over 44 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non perturbed, human eye tissue RNA-seq samples. The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote. The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.02 as of Februay 24 2019, https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues. As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest targets for organoid process improvements. CONCLUSION: The eyeIntegration v1.02 web app serves the pan-eye and body transcriptome dataset, EiaD. This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
---

@@

library(pool)
library(RSQLite)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(dbscan)
library(ggforce)
library(cowplot)
library(viridis)
library(rworldmap)

@@

## Differential Gene Expression Analysis and GO term enrichment
We used the processed data to determine differential gene expression between different sub-tissue types. First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.  We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchielimmaPowersDifferential2015; @lawVoomPrecisionWeights2014]. We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment. The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set was tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].

@@

Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delinerating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue. Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC). Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).

@@

While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regards. We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from the human body tissues. Adult retina is most similar to the brain tissue, pituitary is similar to adult RPE, and skin and blood are similar to cornea.

@@

## Custom gene and tissue expression boxplots
The 'Expression' tab of the webpage provides a wealth of information about both gene- and transcript-level expression for eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`). The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues. A boxplot is then generated after hitting the "Re(Draw) Plot" button with overlaid individual data points. On mouse-over, the metadata for the individual sample is displayed. A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue. The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.

@@

Heatmaps based on expression can be drawn for selected genes and tissues, and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`). Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.

@@

To identify specific changes between organoid and retina tissue, we did differential gene expression and GO term enrichment analyses. GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets. We also noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')`.

We plotted these three sets of genes in the heatmap view, with columns as age in days of fetal or organoid retina and rows are genes. We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`). We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.

@@

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed. Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina development.

Wrapping all of the data processing steps in a Snakemake pipeline has several important features: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with minimal effort, and because all the processing is in modular pieces it is easier to add new analysis steps. In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, adding non-human eye samples, and epigenetic datasets.

@@

```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
rgc_cadherin_hoxb_cap <- supFig_cap(name='rgc_cadherin_hoxb', caption = 'Heatmap of three sets of genes across fetal retina and organoids divided by age in days (A). Bar plot of differential expression from retina to organoid, where positive values are genes that are higher expressed in fetal tissue than organoid. All logFC expression values have FDR corrected p value  < 0.01.')

>>>>

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye tissue RNA-seq samples. The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote. The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (version 1.02 as of Februay 24 2019, https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing users quick visualizations by database version, gene/transcript, and 60 tissues. As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest targets for organoid process improvements. CONCLUSION: The eyeIntegration v1.02 web app serves the pan-eye and body transcriptome dataset, EiaD. This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
---

@@

library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(dbscan)
library(ggforce)
library(cowplot)
library(viridis)
library(rworldmap)

@@

## Differential Gene Expression Analysis and GO term enrichment
We used the processed data to determine differential gene expression between different sub-tissue types. First, we generated a synthetic body set to serve as single representative sub-tissue type for pan-body gene expression by randomly sampling GTEx tissues.  We used the voom function from the limma R package to convert gene expression to precision weights, and then performed pairwise differential expression tests for all combinations of eye sub-tissues, the synthetic body tissue, and human body tissues using an empirical Bayes test [@ritchielimmaPowersDifferential2015; @lawVoomPrecisionWeights2014]. We extracted significant genes (FDR p < 0.01) for all `r limma_DE_tests %>% nrow()` comparisons and used these to calculate GO enrichment. The significant gene list for each eye sub-tissue was split into upregulated and down regulated sets and each set f tested for enrichment using the enrichGO function from the clusterProfiler R package (q-value < 0.01) [@yuClusterProfilerPackageComparing2012].

@@

Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue. Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC). Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).

@@

While t-SNE is a powerful algorithm for grouping samples, it is not consistent for determining the relationships between clusters [@wattenbergHowUseTSNE2016a]; PCA is more useful in this regard. We ran a PCA dimensionality reduction (`r supFig_cap(name = 'allSamplePCA', display = 'cite')`) on all samples, finding that the eye tissues still generally group together and apart from the human body tissues. Adult retina is most similar to the brain tissue, pituitary is similar to adult RPE, and skin and blood are similar to cornea.

@@

## Custom gene and tissue expression boxplots
The 'Expression' tab of the webpage provides a wealth of information about both gene- and transcript-level expression for eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`). The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues. A boxplot is then generated after hitting the "Re(Draw) Plot" button with overlaid individual data points. On mouse-over, the metadata for the individual sample is displayed. A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue. The tables can be arranged or sorted to the user’s preference and can be easily downloaded for local use.

@@

Heatmaps based on expression can be drawn for selected genes and tissues and are useful for comparing gene expression across many genes and tissues (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'B')`). Finally, a session can be saved or shared by building a custom link for the session with the "Build URL Shortcut" button.

@@

To identify specific changes between organoid and retina tissue, we did differential gene expression and GO term enrichment analyses. GO term enrichment identified cell adhesion (protocadherins) and patterning (HOXB family) as enriched gene sets. As there is some evidence suggesting that protocadherins influence RGC viability CITATION and we noticed that several RGC markers appeared to have lower expression in the organoids compared to the fetal tissue `r paste0(fig_cap(name = 'tsneFetalRetina', display = 'cite'),'B')` we looked more closely into RGC marker expression.

We plotted these three sets of genes with a heatmap visualization, with columns as age in days of fetal or organoid retina. Rows are genes, split by the three different groups of genes and are internally clustered by how similar the expression patterns are. We see that there are strong, consistent gene expression changes in these three groups of genes between fetal retina and the organoid samples (`r paste0(fig_cap(name = 'rgc_cadherin_hoxb', display = 'cite'))`). We also plot the differential expression differences between all organoids and all fetal retina samples; all genes across all three sets are significantly differentially expressed with a FDR corrected p value < 0.01.

@@

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for organoid-based models to be developed. Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues, to a first approximation, at the whole transcriptome level. However as organoids do yet develop to full function, it is important to look at how gene expression systems differ between organoid and fetal tissue so as to suggest areas for improvement. https://www.ncbi.nlm.nih.gov/pubmed/12514132

We use our large dataset to narrow in on some processes which to identify three core processes which differ. First we show that the homebox (HOXB) family is overexpressed in the organoids. The homeobox family is well known to initiate polarity of the embryo during early development. As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the application of retinoic acid. While higher HOXB expression may be important to spur differentiation of the multiple cell types of the retina, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.

Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids. Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival. We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result. Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues. This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.

https://www.ncbi.nlm.nih.gov/pubmed/30266927

http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha

http://jcs.biologists.org/content/128/8/1455

We built the Eye in a Disk dataset and the accompanying web app, eyeIntegration in the hopes that easily accessible gene expression across tissue space and time will be a useful tool for hypothesis generation and refinement in eye research. Wrapping all of the data processing steps in a Snakemake pipeline has several important features for the community: our code is publicly available for review, our analyses are reproducible, future sample updates can be streamlined in with less effort, and because all the processing is in modular pieces it is easier to add new analysis steps. In the future, we plan on regularly adding new samples to EiaD, offering *de novo* eye tissue transcriptomes, expanding the single cell RNA-seq expression tooling, adding non-human eye samples, and epigenetic datasets.

@@

```{r rgc_cadherin_hoxb, echo = F, message = F, warning=F, fig.width = 10, fig.pos='H', fig.cap = snakemakeDAG_cap}
rgc_cadherin_hoxb_cap <- supFig_cap(name='rgc_cadherin_hoxb', caption = 'Heatmap of three sets of genes across fetal retina and organoids divided by age in days (A). Bar plot of differential expression from retina to organoid, where positive values are genes that are higher expressed in fetal tissue than organoid (B). All logFC expression values have FDR corrected p value < 0.01.')
